---
title: "Summer Research Project (2009 NFL Season)"
author: "Sujit Sivadanam"
date: "2024-05-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the Dataset

```{r}
#play_by_play <- read.csv("./NFL Play by Play 2009-2017 (v4).csv")
play_by_play <- read.csv("2009_nfl_season.csv")

#play_by_play$Year <- format(as.Date(play_by_play$Date), "%Y")
#write.csv(play_by_play %>% filter(Year == 2017), row.names=F, file="2017_nfl_season.csv")
```

## Necessary Packages

```{r}
library(dplyr)
library(tidyverse)
```

## Store original number of games
```{r}
n <- length(unique(play_by_play$GameID))
```

## Removing certain rows

```{r}
## Sanity check: where Drive numbers are out of order
# drive.not.order <- play_by_play %>% group_by(GameID) %>% summarize(diff = diff(Drive)) %>% summarize(count.bad = sum(diff == -1)) #%>%
#   #filter(diff < 0) #Difference is negative
# length(unique(drive.not.order$GameID))
# 
# drive.not.order
# 
# View(play_by_play %>% filter(GameID %in% drive.not.order$GameID))
# 
# #
# play_by_play %>% group_by(GameID) %>% summarize(diff = diff(Drive)) %>% 
#   filter(diff == -1 & lag(diff) == 0)
# 
# View(play_by_play %>% filter(GameID == 2009091306))

## Define Half variable
play_by_play$Half <- ifelse(play_by_play$qtr %in% c(1,2),
                            1,
                            ifelse(play_by_play$qtr %in% c(3,4),
                                   2,
                                   "OT"))

## where PlayType = "Timeout" or "Two Minute Warning"
play_by_play <- play_by_play %>% 
  filter(!(PlayType %in% c("Timeout", "Two Minute Warning")))

## where PlayType = "Quarter End" and qtr = 1, 3
play_by_play <- play_by_play %>%
  filter(!(PlayType == "Quarter End" & qtr %in% c(1,3)))

unique(play_by_play$PlayType)
```

## Convert Penalty.Yards into signed version (if posteam is penalized, sign is negative) within each drive

```{r}
#check <- play_by_play %>% filter(!is.na(PenalizedTeam) & Penalty.Yards == 0)

play_by_play$Penalty.Yards <- ifelse(is.na(play_by_play$Penalty.Yards),
                                     0, play_by_play$Penalty.Yards)

## Assign sign to Penalty.Yards
play_by_play$Penalty.Yards <- ifelse(play_by_play$posteam ==
        play_by_play$PenalizedTeam, -1*play_by_play$Penalty.Yards,
        play_by_play$Penalty.Yards)
```

## Variable describing field position change (yrdln[current]-yrdln[next]) within each game, each drive

```{r}
play_by_play$pos_net <- c(-1*diff(play_by_play$yrdline100), NA)

## Locates where posteam changes
# which(play_by_play$drive != lead(play_by_play$drive))

play_by_play$pos_net[which(play_by_play$Drive != lead(play_by_play$Drive))] <- NA

## Locate any mismatches between pos_net
mismatch <- which(play_by_play$pos_net != play_by_play$Yards.Gained + play_by_play$Penalty.Yards)

# View profile
View(play_by_play[mismatch,] %>% select(Drive, posteam, yrdln, yrdline100, pos_net, Yards.Gained, Penalty.Yards, Accepted.Penalty, desc, PlayType, qtr))
```

## Create a Kickoff Return.Yards variable

```{r}
play_by_play$Return.Yards <- ifelse(play_by_play$PlayType == "Kickoff",
                                    play_by_play$Yards.Gained, NA)

mismatch_df <- play_by_play[mismatch,] %>% select(Drive, posteam, yrdln, yrdline100, pos_net, Yards.Gained, Return.Yards, Penalty.Yards, Accepted.Penalty, desc, PlayType, qtr) %>% filter(PlayType != "No Play")
```

## Create Fumble.Lost variable
```{r}
play_by_play$Fumble.Lost <- play_by_play$Fumble

play_by_play$Fumble.Lost <- ifelse(play_by_play$Fumble == 1 & play_by_play$Drive != lead(play_by_play$Drive), 1, 0)
```

## Fumble Recovery for TD

```{r}
## All instances of a fumble recovered for touchdown
fmb_td_index <- which(play_by_play$Touchdown == 1 & play_by_play$Fumble == 1 & play_by_play$Fumble.Lost == 0)

## View
View(play_by_play[sort(c(fmb_td_index, fmb_td_index+1, fmb_td_index-1)),] %>% 
       select(posteam, DefensiveTeam, desc, down, PlayType, PosTeamScore, DefTeamScore))

## All instances of fumble
#table(play_by_play$PlayType[play_by_play$Fumble == 1])

## All instances of a touchdown
td_index <- which(play_by_play$Touchdown == 1)
```

## Create Home Team & Away Team Score and Points variables

```{r}
## Home.Team.Sc and Away.Team.Sc
play_by_play$Home.Team.Sc <- ifelse(play_by_play$posteam == play_by_play$HomeTeam, play_by_play$PosTeamScore, play_by_play$DefTeamScore)

play_by_play$Away.Team.Sc <- ifelse(play_by_play$posteam == play_by_play$AwayTeam, play_by_play$PosTeamScore, play_by_play$DefTeamScore)

## Home.Team.Points and Away.Team.Points
play_by_play$Home.Team.Points <- c(diff(play_by_play$Home.Team.Sc), NA)
play_by_play$Away.Team.Points <- c(diff(play_by_play$Away.Team.Sc), NA)

# Makes sure that differences don't include different games
play_by_play$Home.Team.Points[which(play_by_play$GameID != lead(play_by_play$GameID))] <- NA
play_by_play$Away.Team.Points[which(play_by_play$GameID != lead(play_by_play$GameID))] <- NA

## Assign 3 points to home and away teams where their score is NA & field goal made
home_away_NAs <- which(is.na(play_by_play$Home.Team.Points) & play_by_play$FieldGoalResult == "Good")

play_by_play$Home.Team.Points[home_away_NAs] <- ifelse(play_by_play$posteam[home_away_NAs] == play_by_play$HomeTeam[home_away_NAs], 3, 0)

play_by_play$Away.Team.Points[home_away_NAs] <- ifelse(play_by_play$posteam[home_away_NAs] == play_by_play$AwayTeam[home_away_NAs], 3, 0)

## Assign 6 points to home and away teams where their score is NA & TD scored
TD_home_away_NAs  <- which(is.na(play_by_play$Home.Team.Points) &
                             play_by_play$Touchdown == 1)

play_by_play$Home.Team.Points[TD_home_away_NAs] <- ifelse(play_by_play$posteam[TD_home_away_NAs] == play_by_play$HomeTeam[TD_home_away_NAs], 6, 0)

play_by_play$Away.Team.Points[TD_home_away_NAs] <- ifelse(play_by_play$posteam[TD_home_away_NAs] == play_by_play$AwayTeam[TD_home_away_NAs], 6, 0)
# We're yet to know if these TDs are offensive or defensive

## Assign 2 points to home and away teams where their score is NA & safety scored
sfty_NAs <- which(is.na(play_by_play$Home.Team.Points) & play_by_play$Safety == 1)

play_by_play$Home.Team.Points[sfty_NAs] <- ifelse(
  play_by_play$DefensiveTeam[sfty_NAs] == play_by_play$HomeTeam[sfty_NAs], 2, 0)

play_by_play$Away.Team.Points[sfty_NAs] <- ifelse(
  play_by_play$DefensiveTeam[sfty_NAs] == play_by_play$AwayTeam[sfty_NAs], 2, 0)

## Remove the half where TD and Safety == 1
badIndex <- which(play_by_play$Touchdown == 1 & play_by_play$Safety == 1)
play_by_play <- play_by_play %>% filter(!(play_by_play$GameID %in% play_by_play$GameID[badIndex]))
```

## Convert Punting team into Defensive Team

```{r}
## Check for "clean" punts that are NOT of "Punt" playtype
clean.punt <- which(!is.na(play_by_play$PuntResult) & play_by_play$PlayType != "Punt")

## Sanity check to see if drive no. of punt play is different from drive no. of next play
bad.punt.indx <- which(play_by_play$PlayType == "Punt" & play_by_play$Drive == lead(play_by_play$Drive) & !str_detect(tolower(play_by_play$desc), "fake punt"))

## Removes instances of the above
badGameIDs <- play_by_play$GameID[bad.punt.indx]
play_by_play <- play_by_play %>% filter(!(GameID %in% badGameIDs))

## Instances of punt plays
#punt_indx <- which(play_by_play$PlayType == "Punt" & play_by_play$Drive != lead(play_by_play$Drive))

## Punt not followed by 2 consecutive plays of same drive
punt_indx2 <- which(!is.na(play_by_play$PuntResult) & !(play_by_play$Drive == lead(play_by_play$Drive) & play_by_play$Drive == lead(play_by_play$Drive, 2)))

#View(play_by_play[sort(c(punt_indx2-1,punt_indx2,punt_indx2+1)),] %>% select(GameID, qtr, Drive, posteam, DefensiveTeam, desc))


# Switches
play_by_play[punt_indx2, c("posteam", "DefensiveTeam", "PosTeamScore", "DefTeamScore")] <- play_by_play[punt_indx2, c("DefensiveTeam", "posteam", "DefTeamScore", "PosTeamScore")]

## Change drive number into the next drive number
play_by_play$Drive[punt_indx2] <- play_by_play$Drive[punt_indx2]+1

View(play_by_play)
```

## Create Pos.Team.Points variable

```{r}
play_by_play$Pos.Team.Points <- ifelse(play_by_play$posteam == play_by_play$HomeTeam,
                                       ifelse(play_by_play$Home.Team.Points >= play_by_play$Away.Team.Points,
                                              play_by_play$Home.Team.Points,
                                              -1*play_by_play$Away.Team.Points),
                                       ifelse(play_by_play$Away.Team.Points >= play_by_play$Home.Team.Points,
                                              play_by_play$Away.Team.Points,
                                              -1*play_by_play$Home.Team.Points))

## View at instances of touchdowns
View(play_by_play[sort(c(td_index, td_index+1)),] %>%
       select(posteam, DefensiveTeam, desc, down, PlayType, PosTeamScore, DefTeamScore, Home.Team.Sc, Away.Team.Sc, Home.Team.Points, Away.Team.Points, Pos.Team.Points))

## View at instances of safety
View(play_by_play[which(play_by_play$Safety == 1),] %>%
       select(posteam, DefensiveTeam, desc, down, PlayType, PosTeamScore, DefTeamScore, Home.Team.Sc, Away.Team.Sc, Home.Team.Points, Away.Team.Points, Pos.Team.Points))
```

## Extracting last play for each drive of every Game

```{r}
## Alternative way -- Right Here
last.play.drive <- play_by_play %>% group_by(GameID, Drive) %>% summarize(nxt.drive.first.play = head(PlayType,1), nxt.drive.punt.result = head(PuntResult,1), points = tail(Pos.Team.Points,1), qtr = tail(qtr,1), down = tail(down,1), PlayType = tail(PlayType,1), desc = tail(desc,1), Fumble.Lost = tail(Fumble.Lost,1), InterceptionThrown = tail(InterceptionThrown,1), FieldGoalResult = tail(FieldGoalResult,1), ChalReplayResult = tail(ChalReplayResult,1), TwoPointConv = tail(TwoPointConv,1)) %>% ungroup() %>% mutate(nxt.drive.first.play = lead(nxt.drive.first.play), nxt.drive.punt.result = lead(nxt.drive.punt.result), .by="GameID")

## Last play was 6 points without extra point following it
no.xtra.pts <- last.play.drive %>% filter(points == 6)

bad.drive.end <- play_by_play %>% filter(GameID %in% no.xtra.pts$GameID) %>% group_by(GameID) %>% summarize(last.drive = tail(Drive,1)) %>% left_join(no.xtra.pts) %>% 
  filter(last.drive != Drive) %>% .[["GameID"]]

## Removing bad.drive.end (GameID) games from play_by_play
play_by_play <- play_by_play %>% filter(!(GameID %in% bad.drive.end))
```

## Sanity Check

```{r}
# Sanity check: remove extra points not following a touchdown
# bad.ex.pts <- which(play_by_play$PlayType == "Extra Point" & lag(play_by_play$Pos.Team.Points) == -6)
# 
# # Switch posteam and DefTeam at preceding indices and make Pos.Team.Points positive
# play_by_play[bad.ex.pts-1, c("posteam", "DefensiveTeam", "PosTeamScore", "DefTeamScore")] <- play_by_play[bad.ex.pts-1, c("DefensiveTeam", "posteam", "DefTeamScore", "PosTeamScore")]
# 
# play_by_play$Pos.Team.Points[bad.ex.pts-1] <- -1*play_by_play$Pos.Team.Points[bad.ex.pts-1]
# 
# #
# play_by_play %>% group_by(GameID, Drive) %>% 
#   summarize(Drive.Points = sum(Pos.Team.Points)) %>%
#   filter(Drive.Points == 5) #%>% .[["Drive.Points"]] %>% table()
# 
# View(play_by_play %>% filter(GameID == 2009091303 & Drive %in% c(20, 21)))
```

## Drive Outcome for Obvious Cases

```{r}
## Initializing drive outcome
last.play.drive$Drive.Outcome <- NA

## When last.play.points = 3 (field goal)
last.play.drive$Drive.Outcome[last.play.drive$points == 3] <- "Made Field Goal"

## When last.play.points = -2 (safety)
last.play.drive$Drive.Outcome[last.play.drive$points == -2] <- "Safety"

## When last.play.points = 1, 2, or 6 (off TD)
last.play.drive$Drive.Outcome[last.play.drive$points %in% c(1,2,6) | last.play.drive$PlayType == "Extra Point" | !is.na(last.play.drive$TwoPointConv)] <- "Offensive Touchdown"

## When last.play.points == -6 (def TD)
last.play.drive$Drive.Outcome[last.play.drive$points == -6] <- "Defensive Touchdown"
```

## Tackle the 0 points case for Drive.Outcome

```{r}
#View(last.play.drive[last.play.drive$points == 0 & last.play.drive$Fumble == 1,])

## Drive ends with half end
last.play.drive$Drive.Outcome[(last.play.drive$PlayType %in% c("Half End", "End of Game")) | (last.play.drive$PlayType == "Quarter End" & last.play.drive$qtr %in% c(2,4,5))] <- "End of Half"

## Drive ends with interception
last.play.drive$Drive.Outcome[last.play.drive$points == 0 & last.play.drive$InterceptionThrown == 1] <- "Interception"

## Drive ends with fumble
#Change Fumble indicator to 0 due to reversal
fmb_aft <- str_detect(str_remove_all(tolower(last.play.drive$desc), ".*reverse"), "fumble")
last.play.drive$Fumble.Lost[!fmb_aft] <- 0

last.play.drive$Drive.Outcome[last.play.drive$Fumble.Lost == 1] <- "Fumble"

## Drive ends with missed or blocked field goal
last.play.drive$Drive.Outcome[last.play.drive$FieldGoalResult %in% c("No Good", "Blocked")] <- "Missed Field Goal"

## Drive ends with punt
last.play.drive$Drive.Outcome[which(!is.na(last.play.drive$nxt.drive.punt.result))] <- "Punt"

## Remove badly recorded games
bad.record <- last.play.drive$GameID[which(!is.na(last.play.drive$nxt.drive.first.play) & last.play.drive$down != 4 & is.na(last.play.drive$Drive.Outcome))]

play_by_play <- play_by_play %>% filter(!(GameID %in% bad.record))
last.play.drive <- last.play.drive %>% filter(!(GameID %in% bad.record))

## Drive ends with time running out
last.play.drive$Drive.Outcome[which(((is.na(last.play.drive$nxt.drive.first.play) & last.play.drive$down != 4) | (last.play.drive$down == 4 & last.play.drive$PlayType == "Punt")) & is.na(last.play.drive$Drive.Outcome))] <- "End of Half"

## Drive ends with turnover on downs
last.play.drive$Drive.Outcome[is.na(last.play.drive$Drive.Outcome)] <- "Turnover on Downs"
```

## Last Play of Each Game

```{r}
## Extracting last row of each game
last_play <- data.frame()
for(game in unique(play_by_play$GameID)){
  last_row <- tail(subset(play_by_play, GameID == game), n=1)
  last_play <- rbind(last_play, last_row)
}

## Cases of last play of game has Home.Team.Points or Away.Team.Points != 0
View(last_play %>% select(posteam, DefensiveTeam, desc, qtr, down, PlayType, Home.Team.Sc, Away.Team.Sc, Home.Team.Points, Away.Team.Points) %>%
       filter(Home.Team.Points != 0 | Away.Team.Points != 0))
```

## Distribution of Drive Outcomes for each Season

```{r}
table(drive_df$PlayType)
```
